---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

### The `ggformula` and `ggplot` packages

Until now you have used a range of plotting commands to produce scatterplots,
barplots, boxplots etc. that all come from the `ggformula` R package. You may
not have noticed this package, as it is automatically loaded by `bio2020` if
you use it, or by `mosaic`. The `ggformula` package has a very consistent 
syntax, that aligns with most of your data analyses, in that it is:

`gf_plottype(response ~ explanatory variable(s), data = dataset, ...)`

where the `...` are the options. You can read all about using the `ggformula`
package at <http://www.mosaic-web.org/ggformula/articles/pkgdown/ggformula-blog.html>

What you might not be aware of is that `ggformula` actually provides a simple
interface to a much more advanced package called `ggplot2`. The latter has many
powerful features, including creating maps, 3D plots, tree-diagrams, indeed
every sort of graphic you could possibly think of. Hundreds of extensions to
the original `ggplot2` package are now available in R, which allow you to 
produce virtually any plot you could possibly want. You can view a selection of
them at <https://exts.ggplot2.tidyverse.org/gallery/>

Oh, and in case you are wondering why they all seem to have the letters `gg` in
them, it is because they are based on trying to standardise creation of plots
via a "grammar of graphics".

### Transitioning from `ggformula` to `ggplot2`
The first question you might have is _Why do I need to bother learning `ggplot2`?_
Basically, because you might:

* need to create some more complex plots that `ggformula` cannot handle
* you are working with a member of staff (or in your future career) where `ggplot2`
is well understood, whereas `ggformula` is little used outside teaching.

Fortunately, as you already know how to use `ggformula` you will find it very easy
to learn `ggplot2` and this tutorial will teach you the basics.

## Boxplot


```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Histogram

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

